---
title: "Data Tidying - MLB"
author: Greer Moran
output: html_document
date: "Thursday, September 11, 2025"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Front Matter - Clean Environment, Load Libraries, User Defined Functions

Project Uses `Lahman` package from Base R

```{r}
rm(list = ls())

library(tidyverse)
library(dplyr)
library(Lahman)

```

### Pipeline Error Identification and Fix

" mtcars |>
    group_by(cyl) |>
    summarize(avg_mpg = mean(mpg)) |>
    filter(am == 1) "


The filter() verb is nested inside the data set being tidied, with the aim of only taking observations with an entry of 1 for the `am` variable. 
However, a summarize command has already been executed to get the mean of `mpg` for each `cyl` row entry. This means the filter() verb can not be executed past this, unless it related to an observational entry from the `cyl` or `avg_mpg` variable.

To get what is desired, move the filter() verb and it's original argument to be piped before the group_by() function. The output is so,

```{r}
mtcars |>
  filter(am == 1) |>
  group_by(cyl) |>
  summarize(avg_mpg = mean(mpg))
```

### Interpretation of Variable, N
```{r}
mtcars |>
  group_by(cyl) |>
  summarize(
    N = n(),
    avg_mpg = mean(mpg)
  )
```

The data set mtcars gathers data from 32 different cars. `N` is statistically considered the count. So the count in this table is the number of cars recorded by cylinder value. 


### Single Data Verb Function

(a) Find the average of one of the variables. 
USE: summarize()

(b) Add a new column that is the ratio between two variables. 
USE: mutate()

(c) Sort the cases in descending order of a variable. 
USE: arrange()

(d) Create a new data table that includes only those cases that meet a criterion. 
USE: filter()

(e) From a data table with three categorical variables A, B, and C, and a quantitative variable X, produce a data frame that has the same cases but only the variables A and X. 
USE: select()

#### Define batting average (`BA`).

Batting average is the ratio of hits (`H`) to at-bats (`AB`).
```{r}
Teams <- Teams |>
  mutate(BA = H / AB)
```
#### Define slugging percentage (`SLG`). 

Slugging percentage is total bases earned by hits divided by at-bats. 
To compute total bases, get 1 for a single, 2 for a double, 3 for a triple, and 4 for a home run. 

```{r}
Teams <- Teams |>
  mutate(
    X1B = H - (X2B + X3B + HR), 
    totalBases = X1B + X2B*2 + X3B*3 + HR*4,
    SLG = totalBases / AB
    )
```
#### Calculation Verification, Current Findings

```{r}
Teams |>
  select(yearID, teamID, franchID, BA, SLG) |>
  head(10)
```
#### `SLG` (y-axis) as a function of `yearID` (x-axis)
```{r}
ggplot(
  data = Teams,
  mapping = aes(
    x = yearID,
    y = SLG
  )
) + geom_point() +
  geom_smooth(
  method = lm,
  se = FALSE
  ) +
  labs(
    x = "Year",
    y = "Slugging Percentage"
  )
```

#### `BA` as a function of year.
```{r}
ggplot(
  data = Teams,
  mapping = aes(
    x = yearID,
    y = BA
  )
) + geom_point() +
  geom_smooth(
  method = lm,
  se = FALSE,
  col = "red"
  ) +
  labs(
    x = "Year",
    y = "Batting Average"
  )
```


### Finding top teams based on sluggling percentage. 

#### Using the `Teams` data frame in the Lahman package, displayed are the top-5 teams ranked in terms of slugging percentage 

```{r}
Teams |>
  select(yearID, lgID, teamID, franchID, SLG) |>
  arrange(desc(SLG)) |>
  head(5)
```
#### Using teams prior to 2000
```{r}
Teams |>
  filter(yearID < 2000) |>
  select(yearID, lgID, teamID, franchID, SLG) |>
  arrange(desc(SLG)) |>
  head(5)
```

#### `SLG` (y axis) as a function of `yearID` (x-axis) since 1954 conditioned by league (American vs. National, see `lgID`). 

```{r}
plot_Teams <- Teams |>
  filter(yearID >= 1954)

ggplot(
  data = plot_Teams,
  mapping = aes(
    x = yearID,
    y = SLG, 
    shape = lgID,
    color = lgID
  )
) +
  geom_point()+
  geom_smooth(
    method = lm,
    se = FALSE
    ) + 
  scale_color_manual(values = c("red", "skyblue")) +
  labs(
    x = "Year",
    y = "Slugging Percentage", 
    color = "League",
    shape = "League"
  )
```

#### b. Is slugging percentage typically higher in the American League (AL) or the National League (NL)? 

The American League appears to have a slightly higher slugging percentage, on average, compared to the National League.

#### Why there is this pattern in the plot?

The National League makes pitchers hit at bat instead of assigning a designated hitter on the batting roster, hence the DH is used to increase hits. This rule has changed since 2022, however, so this difference in slugging percentage should decrease in years to come. 

